---
title: "Lecture 1: Inroduction"
subtitle: PBIO-03 Design and planning of experiments
author: "Pedro J. Aphalo"
date: "5 September 2017"
output: 
  ioslides_presentation: 
    fig_height: 4
    fig_width: 6
    keep_md: yes
    transition: faster
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
library(ggpmisc)
```

# Aims of design of experiments

## Discussion

- Why is design of experiments a sub-discipline of Statistics?

- What do we aim at when we design an experiment?

- What could be the differences between design and planning of experiments?

- After this talk, you will design some experiments working in groups.

## The aims of design of experiments

**Validity**

- The assumptions needed for a valid data analysis must be fulfilled.

**Efficiency**

- The amount and quality of information obtained with a given effort should be maximized.

**Range of validity**

- The data obtained should be representative for the whole scope of intended use of the acquired information (spatial, temporal, genotypic `space').

## Validity: Assumptions

- Assumptions are never "TRUE", but they can sometimes be close enough to being fulfilled.
- All data analysis procedures depend on some assumptions.
- Some assumptions are fundamental.
- Additional assumptions may be required by certain procedures for data analysis.
- The fewer assumptions made, the more robust are the conclusions obtained.

## Why is bias a problem?

- Bias can be completely confounded with treatment effects.
- It cannot be measured within an experiment and it cannot be diagnosed based on the collected data.
- There is no statistical procedure that can rescue or even identify this problem.
- Bias due to lack of randomization or due to _"subjective randomization"_ or systematic measuring protocols results in _biased (bad) data_ leading to _wrong conclusions_ and **unrepeatable experiments**!

## Avoiding bias

- Randomization of treatment assignment.
- Randomization of measurement sequences.
- Randomization of operator assignment.
- When source of variation is known using restricted randomization (blocks)...
- ... or quantifying the external cause of the variation for each observation (covariates).

## Detecting accidental bias

- Most effective way is to have the whole study replicated independently by other labs.
- We can also go quite far by replication in time by same research group if care is taken so that the same "hidden" disturbances are not likely to repeat.

## Irreproducibility of research results

- Sure ways of compromising reproducibility:
    * consider interdependent observations as replicates ("pseudo-replication"),
    * add replicates until effects are significant,
    * delete apparent outliers until significance is reached ("cherry picking"),
    * choose for publication the "best" replicate from a series of replications of an experiment ("cherry picking"),
    * not reporting methods in detail, including problems encountered,
    * making assumptions that are not tenable or are not validated.
- this is because statistical tests and _p_-values they return are invalidated by bias.

## Reproducibility of research resulst

- Good ways of avoiding gross irreproducibility:
    * avoid "wanting" a certain outcome (not getting in love with an idea).
    * use valid designs and protocols in your experiments.
    * understand the principles of experimental design and always follow them (avoid short cuts).
   
## Real pseudoreplication example

Two ANOVAs of the same anatomical data. Response is stomatal density.
Exciting results with pseudo-replication! (leaves as replicates)

|Effect |	Df  |	Sum Sq	 | Mean Sq| F value |	Pr(>F)|
|:------|----:|---------:|--------:|--------:|-----:|
filter	| 5 |	28900	| 5773 | 3.14 |	0.0090 |
block	| 2	| 24600	| 12292	| 6.69	| 0.0015 |
Residuals	| 208	| 381914 | 1836	| |

## Real pseudoreplication example

Two ANOVAs of the same anatomical data. Response is stomatal density.
Boring results with valid analysis! (plots as replicates)

| Effect | Df | Sum Sq | Mean Sq | F value | Pr(>F) |
|:------|----:|---------:|--------:|--------:|-----:|
| filter | 5 | 1560 | 312 | 0.395 | 0.84 |
| block | 2 | 3340 | 1668 | 2.106 | 0.17 |
| Residuals | 10 | 7920 | 792	 |  |

