---
title: "The light environment"
author: "Pedro J. Aphalo"
format: html
server: shiny
---

```{r, message=FALSE}
library(lubridate)
library(ggplot2)
library(patchwork)
library(ggpmisc)
library(ggspectra)
library(photobiologyPlants)
library(photobiologySun)
library(photobiologyLamps)
library(photobiologyLEDs)
```

## Quantities and units for light

```{r}
inputPanel(
  selectInput("quantity", label = "Spectrum expressed as:",
              choices = c("energy irradiance" = "e_irrad", 
                          "photon irradiance" = "q_irrad",
                          "illuminance" = "illuminance",
                          "PAR photon irraddiance (PPFD)" = "PAR_q_irrad",
                          "PAR energy irraddiance" = "PAR_e_irrad",
                          "UV GPAS (Caldwell's)" = "GEN_G_e_irrad")),
  selectInput("light_source", label = "Light source:",
              choices = c("sun", "cloudy", "vegetation shade", 
                          "neutral white LED", "cool white LED", "warm white LED",
                          "grow light broad spectrum", "R+B LED grow light")),
  sliderInput(inputId = "wl.range0", label = "Wavelength range (nm)", min = 250, max = 900, value = c(280, 800)),
  
)
plotOutput("spctPlot")
```

```{r, message=FALSE}
#| context: server
library(ggspectra)
library(photobiologyPlants)
library(photobiologySun)
library(photobiologyLamps)
library(photobiologyLEDs)

output$spctPlot <- renderPlot({
    spct <- switch(input[["light_source"]],
                   sun = sun.spct,
                   cloudy = sun.spct,
                   "vegetation shade" = sun.spct,
                   "neutral white LED" = lamps.mspct$Airam.LED.14W.4000K,
                   "cool white LED" = lamps.mspct$V.light.LED.2W.6000K,
                   "warm white LED" = lamps.mspct$Osram.LED.10W.2700K,
                   "grow light broad spectrum" = lamps.mspct$Valoya.LED.B50.AP67.LED.grow.lamp,
                   "R+B LED grow light" = leds.mspct$Epistar_3W_Plant_Grow_LED)
    if (grepl("PAR", input[["quantity"]])) {
      w.band <- PAR()
      plot.w.band <- PAR()
    } else if (grepl("GEN_G", input[["quantity"]])) {
      w.band <- GEN_G()
      plot.w.band <- PAR()
    } else if (input[["quantity"]] == "illuminance") {
      w.band <- ciev2.spct
      plot.w.band <- VIS_bands()
    } else {
      w.band <- waveband(spct)
      plot.w.band <- Plant_bands(std = "sensory10")
    }
    if (grepl("e_", input[["quantity"]])) energy_as_default() else photon_as_default()
    
    z <- spct * w.band
    if (is.waveband(w.band)) {
      z <- interpolate_spct(z, spct$w.length[-nrow(spct)], fill = 0)
    }
    
  autoplot(z, geom = "spct", w.band = plot.w.band,
           range = input[["wl.range0"]], span = 51) +
    theme_bw(15)
})
```

## PAR in energy or photon units

The definition of PPFD or PAR photon irradiance of McCree assumes equal effectiveness per photon. How does the definition of PAR look like expressed in an energy basis?

```{r}
inputPanel(
  selectInput("unit_out", label = "PAR response on:",
              choices = c("photon basis " = "photon",
                          "energy basis " = "energy")),
)
plotOutput("parPlot")
```

```{r, message=FALSE}
#| context: server
library(ggspectra)
library(photobiologyWavebands)

output$parPlot <- renderPlot({
  autoplot(PAR(), unit.in = "photon", unit.out = input[["unit_out"]], geom = "spct") +
    theme_bw(15)
})
```

## Sunlight through a day

PAR photon irradiance we all know varies through the day. How does the shape of the spectrum vary? We can normalise the spectral irradiance from different times of the day to be 1 at, for example, 550 nm. We can alternatively re-scale the spectra to, for example, 100 umol m^-2^ s^-1^ of PAR.

```{r}
inputPanel(
  selectInput("unit_out1", label = "Spectrum expressed as:",
              choices = c("photon irradiance" = "photon",
                          "energy irradiance" = "energy")),
  # selectInput("season", label = "Season in Helsinki:",
  #             choices = c("summer",
  #                         "winter")),
  selectInput("norm1", label = "Normalised/scaled: ",
              choices = c("no", "1 at 550 nm" = "wl550", "PAR 100 umol m-2 s-1 " = "PAR100")),
  sliderInput(inputId = "wl.range1", label = "Wavelength range (nm)", min = 250, max = 900, value = c(280, 800), post = " nm"),
  
  sliderInput(inputId = "spct_idx", label = "Time of day (sequence)", min = 1, max = 18, value = 11, step = 1)
)
plotOutput("dayPlot")
```

```{r, warning=FALSE, message=FALSE}
#| context: server
library(ggspectra)
library(photobiologySun)

sun_hourly_june.mspct <- subset2mspct(sun_hourly_june.spct)[1:18]
sun_hourly_june_norm550.mspct <- normalise(sun_hourly_june.mspct, norm = 550)
sun_hourly_june_PAR1000.mspct <- fscale(sun_hourly_june.mspct, range = PAR(), f = q_irrad, target = 100)
mspct.names <- names(sun_hourly_june.mspct)
output$dayPlot <- renderPlot({
  i <- input[["spct_idx"]]
  mspct <- switch(input[["norm1"]],
                 wl550 = sun_hourly_june_norm550.mspct,
                 PAR100 = sun_hourly_june_PAR1000.mspct,
                 no = sun_hourly_june.mspct)
    autoplot(mspct[c(11, i)], 
           range = input[["wl.range1"]],
           unit.out = input[["unit_out1"]], geom = "spct", norm = "skip",
           ylim = ifelse(input[["norm"]] == "max", c(NA, NA), c(NA, 6))) +
#    ggtitle(mspct.names[1]) +
    theme_bw(15) + theme(legend.position = "top")
})
```

## The daily pattern at 1 h time step

Same data as above, simulated hourly means.

```{r}
sun_hourly_june.mspct <- subset2mspct(sun_hourly_june.spct)[1:18]
q_irrad.tb <- q_irrad(sun_hourly_june.mspct, w.band = Plant_bands("CIE"))
q_irrad.tb$time <- ymd_hms(as.character(q_irrad.tb$spct.idx))

(ggplot(q_irrad.tb, aes(time, Q_PAR)) +
  geom_spct() + geom_line() +
    theme_bw()) /
  (ggplot(q_irrad.tb, aes(time, `Q_]UVB.CIE`)) +
  geom_spct() + geom_line() +
    theme_bw(15))
```

## The daily pattern at 1 min time step

Mid summer in Helsinki, measurements are means for each minute compute from measurements once every 5 seconds. The line displays 4440 observations and zooming into the time scale makes visible short-term variability. The 23 June 2022 was a day with broken clouds. The instrumentation of the weather station is described at its [web site](https://viikki-stn.r4photobiology.info/).

```{r}
inputPanel(
  selectInput("waveband", label = "Region of the spectrum:",
              choices = c("PAR", 
                          "Red light",
                          "Far-red light",
                          "Blue light",
                          "UV-A radiation",
                          "UV-A1 radiation",
                          "UV-A2 radiation",
                          "UV-B radiation")),
  sliderInput(inputId = "time.range", label = "Local solar time (hh)", min = 0, max = 24, value = c(0, 24)),
  
)
plotOutput("daycoursePlot")
```

```{r, message=FALSE}
#| context: server
library(rlang)
library(ggplot2)
library(photobiology)

load("one-day-tb.Rda")

layers_scales <- function(my.fill = "grey50", tod.limits = c(0, 24)) {
  list(geom_spct(fill = my.fill, alpha = 0.3),
       geom_line(),
       scale_x_continuous(name = "Local solar time (hh) at Helsinki, 23 June 2022", 
                          breaks = c(0, 6, 12, 18, 24),
                          limits = tod.limits))
}

output$daycoursePlot <- renderPlot({
    qty.colname <- switch(input[["waveband"]],
                   PAR = "PAR_umol_BF",
                   "Red light" = "red_umol",
                   "Far-red light" = "far_red_umol",
                   "Blue light" = "blue_umol",
                   "UV-A radiation" = "UVA_umol",
                   "UV-A1 radiation" = "UVA1_umol",
                   "UV-A2 radiation" = "UVA2_umol",
                   "UV-B radiation" = "UVB_umol")
  qty.colours <- c(PAR = "yellow",
                   "Red light" = "red",
                   "Far-red light" = "darkred",
                   "Blue light" = "cornflowerblue",
                   "UV-A radiation" = "thistle4",
                   "UV-A1 radiation" = "blue",
                   "UV-A2 radiation" = "purple",
                   "UV-B radiation" = "violet")
    
  ggplot(one_day.tb[ , c("solar_time", qty.colname)],
         aes(solar_time, .data[[qty.colname]])) +
  layers_scales(qty.colours[input[["waveband"]]], input[["time.range"]])  +
  labs(y = expression("Photon irradiance  "*(mu*mol~s^{-1}~m^{-2}))) +
    labs(title = input[["waveband"]]) +
    theme_bw(17)

})
```
