---
title: "From lecture 1"
subtitle: Design and planning of experiments
author: "Pedro J. Aphalo"
date: "5 September 2017"
output: 
  ioslides_presentation: 
    fig_height: 4
    fig_width: 6
    keep_md: yes
    transition: faster
    widescreen: yes
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
library(ggpmisc)
```

# Demos

## Introduction

- In the course we will use some demonstrations that you will run interactively through a web browser interface. 
- The first one is for you to play with simulated experiments.
- Data for a _Control_ and a _Treatment_ are simulated by drawing random samples from the Normal distribution.
- The mean of the Control distribution is kept at zero.
- You can choose the value of the mean for Treatment.
- You can also choose the value of the standard deviation used.
- The plot shows: simualted observations in black, sample means and standard errors in red.
- A table with results from ANOVA on the data.

## Simulation experiment: data

```{r sampling, cache=FALSE}
inputPanel(
  selectInput("mean_treat", label = "Mean of 'Treatment':",
              choices = c(0.001, 0.01, 0.1, 1, 10, 100), selected = 10),
  
  selectInput("n_replicates", label = "Number of replicates:",
              choices = c(3, 5, 10, 20, 40, 100, 1000, 10000), selected = 100),
  
  sliderInput("sd_adjust", label = "Standard deviation:",
              min = 0, max = 10, value = 1, step = 0.2),
  
  actionButton("refresh", "New simulation")
  
)

generate_samples <- function() {
  num_replicates <- as.numeric(input$n_replicates)
  num_mean <- as.numeric(input$mean_treat)
  y <- c(rnorm(n = num_replicates, mean = 0, sd = input$sd_adjust),
         rnorm(n = num_replicates, mean = num_mean, sd = input$sd_adjust))
  group <- factor(rep(c("Control", "Treatment"), rep(num_replicates, 2)))
  data.frame(y, group)
} 

renderPlot({
  ggplot(generate_samples(), aes(y, colour = group)) +
    geom_vline(xintercept = c(0, as.numeric(input$mean_treat)), 
               linetype = "dotted", color = "blue") +
    geom_rug() +
    geom_density(fill = NA) +
    annotate(x = 1.5, y = 0, label = input$refresh, alpha = 0, geom = "text") +
    theme_bw(20)
})
```

## Exercise

- Keep constant the standard deviation of the populations sampled.
    * change the Treatment effect size.
    * what happens to the distributions?
- Keep the Treatment effect size constant.
    * change the standard deviation of the populations sampled.
    * what happens to the distributions?
- Keep the Treatment effect size and the standard deviation constant.
    * change number of replicates and see how this affects the empirical density distribution.
    * for each condition run 15 or more simulations, and look at how much variation there is between simulations.

## Simulation experiment: summaries

```{r summaries, cache=FALSE}
inputPanel(
  selectInput("mean_treat1", label = "Mean of 'Treatment':",
              choices = c(0.001, 0.01, 0.1, 1, 10, 100), selected = 1),
  
  selectInput("n_replicates1", label = "Number of replicates:",
              choices = c(3, 5, 10, 20, 40, 100, 1000, 10000), selected = 10),
  
  sliderInput("sd_adjust1", label = "Standard deviation:",
              min = 0, max = 10, value = 1, step = 0.2),
  
  actionButton("refresh1", "New simulation")
  
)

renderPrint({
  num_replicates <- as.numeric(input$n_replicates1)
  num_mean <- as.numeric(input$mean_treat1)
  y <- c(rnorm(n = num_replicates, mean = 0, sd = input$sd_adjust1),
         rnorm(n = num_replicates, mean = num_mean, sd = input$sd_adjust1))
  group <- factor(rep(c("Control", "Treatment"), rep(num_replicates, 2)))
  df <- data.frame(y, group)
  data.frame(group = c("Control", "Treatment"),
             n = num_replicates,
             pop.mean = c(0, num_mean),
             samp.mean = signif(c(mean(df[df$group == "Control", "y"]),
                           mean(df[df$group == "Treatment", "y"])), 3),
             pop.sd = input$sd_adjust1,
             samp.sd = signif(c(sd(df[df$group == "Control", "y"]),
                         sd(df[df$group == "Treatment", "y"])), 3),
             simulation = rep(input$refresh1[1], 2)
  )
})

```

## Exercise

- Keep constant the standard deviation of the populations sampled.
    * change the Treatment effect size.
    * what happens to the mean and sd?
- Keep the Treatment effect size constant.
    * change the standard deviation of the populations sampled.
    * what happens to the mean and sd?
- Keep the Treatment effect size and the standard deviation constant.
    * change number of replicates and see how this affects the sample mean and sd.
    * for each condition run 15 or more simulations, and look at how much variation there is between simulations.

## Simulation experiment: analysis

```{r replication, cache=FALSE}
inputPanel(
  selectInput("mean_treat2", label = "Mean of 'Treatment':",
              choices = c(0.001, 0.01, 0.1, 1, 10, 100), selected = 1),
  
  selectInput("n_replicates2", label = "Number of replicates:",
              choices = c(3, 5, 10, 20, 40, 100, 1000, 10000), selected = 10),
  
  sliderInput("sd_adjust2", label = "Standard deviation:",
              min = 0, max = 10, value = 1, step = 0.2),
  
  actionButton("refresh2", "New simulation")
  
)

generate_data <- function() {
  num_replicates <- as.numeric(input$n_replicates2)
  num_mean <- as.numeric(input$mean_treat2)
  y <- c(rnorm(n = num_replicates, mean = 0, sd = input$sd_adjust2),
         rnorm(n = num_replicates, mean = num_mean, sd = input$sd_adjust2))
  x <- factor(rep(c("Control", "Treatment"), rep(num_replicates, 2)))
  data.frame(y, x)
} 

renderPlot({
  ggplot(generate_data(), aes(x, y)) +
    geom_point(alpha = 0.25) +
    stat_summary(fun.data = "mean_se", color = "red") +
    geom_hline(yintercept = c(0, as.numeric(input$mean_treat2)), 
               linetype = "dotted", color = "blue") +
    stat_fit_tb(tb.type = "anova", size = 3.7) +
    annotate(x = 1.5, y = 0, label = input$refresh2, alpha = 0, geom = "text") +
    theme_bw(20)
})

```

## Exercise

- Keep the Treatment effect size constant.
    * change number of replicates and see how this affects the _p_-value and the plotted standard error.
    * does a non-significant _p_-value demonstrate lack of treatment effect?
- Keep the Treatment effect size constant.
    * change the standard deviation of the populations sampled.
    * does a non-significant _p_-value demonstrate lack of treatment effect?
- What can you conclude from a significant _p_-value?
- Can you conclude something more than "significant effect"?
